---
title: "appendix"
format: html
---

```{r}
# Load packages
library(tidycensus)
library(tidyverse)
library(knitr)
library(scales)
library(RColorBrewer)
library(sf)
library(ggplot2)
library(tigris)
library(patchwork)
library(here)
```

# PHASE 1: DATA PREPARATION

## 1.1 Load and clean Philadelphia sales data
```{r}
# Load Philly Property Sales data
phl_sales <- read_csv("../data/raw/opa_properties_public.csv", show_col_types = FALSE)
```

### 1.1a Filter to residential properties, 2023-2024 sales
```{r}
# Check data types
# glimpse(phl_sales)

phl_sales_res_23_24 <- phl_sales |>
  filter(
    category_code == 1, # Residential
    year(sale_date) %in% c(2023, 2024), # 2023-24 sales
    !is.na(category_code) & !is.na(sale_date) # Handle nulls
  )
```

### 1.1b Remove obvious errors
```{r}
phl_sales_res_23_24 <- phl_sales_res_23_24 |>
  filter(
    # Some sale_price are unrealistically too low ($0, $1 etc.)
    sale_price >= 10000,
    # Some areas are unrealistically low (0, 1, etc.)
    total_area > 1,
    # Some 0's remain in total_liveable_area after first area filter
    total_livable_area > 0
    # MORE: some areas seem unrealistically large (maybe bound by < 25K)
    )
```

### 1.1c Handle missing values

### 1.1d Document all cleaning decisions

- Filter for only residential properties & sales made in 2023-24 (per instructions).


- Filter for realistic sales price values; boundary > $10,000


- Filter for realistic house total area: boundary > 1 sq ft


- 


